Visualizing Independence Using Extended Association Plots

### 2003-08-08 trunk/vcd/paper/dsc2003.Rnw
small bug fixes & final paper update for DSC paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@276 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[10pt]{article}
\usepackage{dsc2003}
\usepackage{Sweave, graphicx}
\usepackage{Z}

\setlength{\parskip}{0.5ex plus0.1ex minus0.1ex}
\setlength{\parindent}{0em}

\bibpunct{(}{)}{;}{a}{}{,}

@@

\SweaveOpts{engine=R}
\setkeys{Gin}{width=0.8\textwidth}

<<preliminaries,echo=FALSE,results=hide>>=
set.seed(12345)
library(vcd)
data(Bundesliga)
goals <- xtabs(~ AwayGoals + HomeGoals, data = Bundesliga, subset = Year == 1995)
data(HairEyeColor)
haireye <- xtabs(Freq ~ Hair + Eye, data = as.data.frame(HairEyeColor), subset = Sex == "Female")
data(UCBAdmissions)
@

\title{Visualizing Independence Using Extended Association Plots}
\author{\hfill David Meyer$^\dag$ \hfill Achim Zeileis$^\dag$ \hfill Kurt Hornik$^\ddag$ \hfill \hfill \\[0.2cm]
\dag {\it \small Institut f\"ur Statistik \& Wahrscheinlichkeitstheorie,
 Technische Universit\"at Wien}\\
\ddag {\it \small Institut f\"ur Statistik, Wirtschaftsuniversit\"at Wien}
}
\date{}
\maketitle

\begin{abstract}
Association plots---a visualization technique for the independence problem in 2-way
contingency tables---are extended in three directions:
\begin{enumerate}
 \item The visualization is enhanced by using colors for the
  importance of the residuals.
 \item Two methods for the use of multi-way tables are proposed.
 \item The implementation currently available in \textsf{R} is improved using a more modular
  design and allowing a more flexible specification of plotting parameters.
\end{enumerate}
\end{abstract}

@@

Given two categorical variables, or, equivalently, their contingency
table, one is often interested in whether they are 
independent or not. Usually, the $\chi^2$ test statistic---the sum of the
squared Pearson residuals---is used as a measure of association.
Now, whenever the statistic is
significant, one might ask for a more detailed analysis on the basis
of the residuals themselves.
This task can greatly be sustained by the use of graphical
methods such as the association plot, for which we propose some
extensions: the use of colors for the shading of the residuals, two
possible methods for multi-way tables, and a new, more flexible
implementation using {\sf R}'s alternative graphical system
\texttt{grid} \citep{vcd:Murrell:2002}. The new functionality is provided 
by the {\sf R} package
\texttt{vcd}, inspired by the book `Visualizing Categorical Data'
\citep{vcd:Friendly:2000}.

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[htbp]
\begin{center}
<<assoc-improved1,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.assocplot(HairEyeColor[,,2], gp = gp.shading)
@
\caption{Extended association plot for `Hair' and `Eye' colors of female students.}
\label{fig:assoc-improved1}
\end{center}
\end{figure}
\pagebreak
\section{Improved visualization through residual shading}

We propose to use a color-shading similar to the
enhancements of \cite{vcd:Friendly:1994} for the mosaic plot 
\cite[a graphical tool for the 
visualization of the observed frequencies in a contingency table; 
see][]{vcd:Hartigan+Kleiner:1984}:
this extension 
uses a color coding of the mosaic tiles to visualize deviations
(residuals) from a given log-linear model fitted to the table.
Positive and negative signs of the residuals are coded by rectangles with
solid and dashed borders respectively. Furthermore, residuals can be
classified according to specified shading levels: per default,
residuals exceeding an absolute
value of 2 are shaded light blue and red respectively, those that even
exceed an absolute value of 4 are shaded with full saturation. The heuristic
behind this shading is that
the Pearson residuals are approximately standard normal which implies
that the highlighted cells are those with
residuals \emph{individually} significant at approximately the 5\% and 0.01\%
level. But the main purpose of the shading is not to visualize
significance but the \emph{pattern} of
deviation from independence \citep[p.~109]{vcd:Friendly:2000}.

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[htbp]
\begin{center}
<<assoc-improved,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.assocplot(t(margin.table(HairEyeColor, c(1,3))), gp = gp.shading)
@
\caption{Extended association plot for `Hair' and `Sex' in the `HairEyeColor' data.}
\label{fig:assoc-improved2}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
<<UCB-mosaic-signif,fig=TRUE,echo=FALSE,height=8,width=8>>=
 panel.assocplot <- function (x, i, j, ..., axes, abbreviate) grid.assocplot(margin.table(x, c(i,j)), ..., panel = T, axis.labels = F, labels = F, main = NULL) 
grid.mosaicpairs(aperm(HairEyeColor), panel.upper= panel.assocplot, panel.lower = panel.assocplot, gp = gp.shading, main = NULL)
@
\caption{Extended association pairs for the `HairEyeColor' data,
  visualizing all pairwise marginal associations.}
\label{fig:HEC-pairs}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
<<UCB-assoc-signif,fig=TRUE,echo=FALSE,height=10,width=10>>=
tabplot(UCBAdmissions, gp = gp.shading)
@
\caption{Extended conditioning association plots for the `UCB Admissions' data.}
\label{fig:UCB-assoc-signif}
\end{center}
\end{figure}

The current implementation of association plots in \textsf{R} suffers
from two main disadvantages: First, it is not easy to recycle the
plots in conditioning plots or pairs plots as
they have been implemented using \textsf{R}'s base graphics engine where in
general plotting to relative coordinates is not supported.
The new implementation was written from scratch in \texttt{grid} offering
much more versatility amongst some minor advantages and convenient
improvements. Second, the graphics parameters of the rectangles in the
association plots, such as color and line type, cannot be specified for each cell
by the user. To overcome this, the current implementation in \texttt{vcd}
allows the user to specify either arrays of graphics parameters
of the same dimensionality as the object being plotted or a function
which computes these graphics parameters based on the original table
and its Pearson residuals. Functions are provided for shading schemes
like those described in the previous sections.

\section{Conclusion}
We suggest a set of enhancements for visualizing the independence
problem in contingency tables using association plots. The
extensions aim at improving the visualization by displaying
the size of the residuals, and by extending the concept of association
plots to multi-way tables. Furthermore, a new
implementation is outlined based on the graphics package \texttt{grid}
which provides more modular design and more flexible specification
of graphical parameters.
Our work is still not at its end, though: for example, the plots for
multi-way tables currently do not use the same scale for all subplots,
and thus residuals of different subplots are not
comparable. Furthermore, the current color scheme only allows the
detection of ``patterns'' of independence. It would be nice, however, 
if the shading levels were
chosen in a way that a true visual test of independence can be performed,
that is, that the presence of ``significant'' residuals leads to a
rejection of the hypothesis of independence. Current work includes the
development of such color schemes, based on an alternative test
statistic---the maximum of the absolute values of the Pearson residuals.

\bibliographystyle{asa}
\bibliography{vcd}

\section*{Addresses}

David Meyer\\
Achim Zeileis\\
E-mail: \texttt{\{David.Meyer,Achim.Zeileis\}@ci.tuwien.ac.at}\\
Institut f\"ur Statistik und Wahrscheinlichkeitstheorie\\
Technische Universit\"at Wien\\
Wiedner Hauptstra{\ss}e 8--10/1071\\
1040 Wien\\
\\
Kurt Hornik\\
E-mail: \texttt{Kurt.Hornik@wu-wien.ac.at}\\
Institut f\"ur Statistik\\
Wirtschaftsuniversit\"at Wien\\
Augasse 2--6\\
1090 Wien\\

\end{document}
% LocalWords:  proofness

>>>>

\documentclass[10pt,a4paper]{article}
\usepackage{dsc2003}
\usepackage[authoryear,round,longnamesfirst]{natbib}
%% don't need: \usepackage{vcd}

@@

\SweaveOpts{engine=R,eps=FALSE}
\setkeys{Gin}{width=0.8\textwidth}

<<preliminaries,echo=FALSE,results=hide>>=
set.seed(12345)
library(vcd)
data(Bundesliga)
goals <- xtabs(~ AwayGoals + HomeGoals, data = Bundesliga, subset = Year == 1995)
data(HairEyeColor)
haireye <- xtabs(Freq ~ Hair + Eye, data = as.data.frame(HairEyeColor), subset = Sex == "Female")
data(UCBAdmissions)

gp.old <- function(x, y) gp.Friendly(x, y, lty = 1)
@

\title{Visualizing Independence Using Extended Association Plots}
\author{David Meyer \And Achim Zeileis \And Kurt Hornik}
\date{}
\maketitle

\begin{abstract}
Association plots---a visualization technique for the independence problem in 2-way
contingency tables---are extended in three directions:
\begin{enumerate}
 \item The visualization is enhanced by using colors for the
  importance of the residuals.
 \item The implementation currently available in \textsf{R} is improved using a more modular
  design and allowing a more flexible specification of plotting parameters.
 \item Two methods for the use of multi-way tables are proposed.
\end{enumerate}
\end{abstract}

@@

Given two categorical variables, or, equivalently, their contingency
table, one is often interested in whether they are
independent or not. Usually, the $\chi^2$ test statistic---the sum of the
squared Pearson residuals---is used as a measure of association.
Now, whenever the statistic is
significant, one might ask for a more detailed analysis on the basis
of the residuals themselves.
This task can greatly be sustained by the use of graphical
methods such as the mosaic display
\citep{vcd:Hartigan+Kleiner:1981,vcd:Friendly:1994,vcd:Theus+Lauer:1999,vcd:Hofmann:2003}
or the association plot \citep{vcd:Cohen:1980,vcd:Friendly:2000}.
For the latter we propose some
extensions: the use of colors for the shading of the residuals, two
possible methods for multi-way tables, and a new, more flexible
implementation using {\sf R}'s alternative graphical system
\texttt{grid} \citep{vcd:Murrell:2002}. The new functionality is provided
by the {\sf R} package
\texttt{vcd}, inspired by the book `Visualizing Categorical Data'
\citep{vcd:Friendly:2000}.

\pagebreak

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[htbp]
\begin{center}
<<assoc-improved1,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.assocplot(HairEyeColor[,,2], gp = gp.old)
@
\caption{Extended association plot for `Hair' and `Eye' colors of female students.}
\label{fig:assoc-improved1}
\end{center}
\end{figure}
\pagebreak
\section{Improved visualization through residual shading}

We propose to use a color-shading similar to the
enhancements of \cite{vcd:Friendly:1994} for the mosaic plot
\cite[a graphical tool for the
visualization of the observed frequencies in a contingency table;
see][]{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984}.
This extension uses a color coding of the mosaic tiles to visualize deviations
(residuals) from independence. The sign and size of residuals are
coded by color shadings of the corresponding rectangles: by default,
residuals with an absolute value smaller than 2 are white, positive
and negative residuals exceeding an absolute
value of 2 are shaded light blue and red respectively, and those that even
exceed an absolute value of 4 are shaded with full saturation. The heuristic
behind this shading is that
the Pearson residuals are approximately standard normal which implies
that the highlighted cells are those with
residuals \emph{individually} significant at approximately the 5\% and 0.01\%
level. But the main purpose of the shading is not to visualize
significance but the \emph{pattern} of
deviation from independence \citep[p.~109]{vcd:Friendly:2000}.

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[htbp]
\begin{center}
<<assoc-improved,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.assocplot(t(margin.table(HairEyeColor, c(1,3))), gp = gp.old)
@
\caption{Extended association plot for `Hair' and `Sex' in the `HairEyeColor' data.}
\label{fig:assoc-improved2}
\end{center}
\end{figure}

The current implementation of association plots in \textsf{R} suffers
from two main disadvantages: First, it is not easy to recycle the
plots in conditioning plots or pairs plots as they have been implemented
using \textsf{R}'s base graphics engine where in general plotting to relative
coordinates is not supported. Second, the graphics parameters of the rectangles in the
association plots, such as color and line type, cannot be specified for each cell
by the user. To overcome this, a new implementation was written from scratch
in \texttt{grid} offering much more versatility amongst some minor advantages
and convenient improvements. It allows the user to specify either 2-way tables of graphics parameters,
i.e., arrays of the same dimensionality as the object being plotted, or a function
which computes these graphics parameters based on the original table
and its Pearson residuals. This makes it easy for users to implement their
own shading schemes; \texttt{vcd} provides functions for shading schemes
like those described in the previous sections. How association plots can
be recycled using the new \texttt{grid} implementation is described in the next
section.

@@

\begin{figure}[htbp]
\begin{center}
<<UCB-mosaic-signif,fig=TRUE,echo=FALSE,height=8,width=8>>=
panel.assocplot <- function (x, i, j, ..., axes, abbreviate)
  grid.assocplot(t(margin.table(x, c(i,j))), ..., panel = TRUE, axis.labels = FALSE, labels = FALSE, main = NULL)
grid.mosaicpairs(aperm(HairEyeColor), panel.upper= panel.assocplot, panel.lower = panel.assocplot, gp = gp.old, main = NULL)
@
\caption{Extended association pairs for the `HairEyeColor' data,
  visualizing all pairwise marginal associations.}
\label{fig:HEC-pairs}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
<<UCB-assoc-signif,fig=TRUE,echo=FALSE,height=10,width=10>>=
tabplot(UCBAdmissions, gp = gp.old)
@
\caption{Extended conditioning association plots for the `UCB Admissions' data.}
\label{fig:UCB-assoc-signif}
\end{center}
\end{figure}

\section{Conclusion}
We suggest a set of enhancements for visualizing the independence
problem in contingency tables using association plots. The
extensions aim at improving (a) the visualization by displaying
the size of the residuals and (b) the implementation based on the
graphics package \texttt{grid} which provides more modular design
and more flexible specification of graphical parameters. Furthermore,
some extensions of association plots to multi-way tables are outlined.
Our work is still not at its end, though: for example, the plots for
multi-way tables currently do not use the same scale for all subplots,
and thus residuals of different subplots are not
comparable. Furthermore, the current color scheme only allows the
detection of ``patterns'' of independence. It would be nice, however,
if the shading levels were
chosen in a way that a true visual test of independence can be performed,
that is, that the presence of ``significant'' residuals leads to a
rejection of the hypothesis of independence. Current work includes the
development of such color schemes, based on an alternative test
statistic---the maximum of the absolute values of the Pearson residuals.

\bibliographystyle{abbrvnat}
\bibliography{vcd}

\subsection*{Affiliation}

David Meyer, Achim Zeileis\\
Institut f\"ur Statistik und Wahrscheinlichkeitstheorie\\
Technische Universit\"at Wien\\
Wiedner Hauptstra{\ss}e 8--10/1071\\
A-1040 Wien, Austria\\
E-mail: \email{David.Meyer@ci.tuwien.ac.at}, \email{Achim.Zeileis@ci.tuwien.ac.at}\\
\\
Kurt Hornik\\
Institut f\"ur Statistik\\
Wirtschaftsuniversit\"at Wien\\
Augasse 2--6\\
A-1090 Wien, Austria\\
E-mail: \email{Kurt.Hornik@wu-wien.ac.at}\\

\end{document}
